---
title: "PM 592 Final Project"
author: "Flemming Wu"
date: "`r Sys.Date()`"
---

```{r clear env and read in data, message=F, warning=F}
library(arrow)
library(tidyverse)
library(skimr)
library(gt)
library(mfp)
library(sjPlot)
library(gtools)
library(psych)
library(gridExtra)
library(ggfortify)
library(olsrr)


rm(list = ls())

nhanes <- arrow::read_parquet("./data/nhanes.parquet")
```

```{r}
psych::describe(nhanes)
```

```{r table 1}
psych::describe(nhanes) %>%
  select(-c(vars, trimmed, mad, range, skew, kurtosis, se)) %>%
  slice(-c(1, 2)) %>%
  cbind(variables = c("Glucose (mg/dL)", "Gender (1 = Male, 2 = Female)", "Age", "BMI (kg/m^2)", "Alcohol Frequency", "Diet Health Score", "Sleep Hours (Weekdays or Weekends)", "MET (activity score)")) %>% 
  relocate(variables) %>%
  sjPlot::tab_df(
    x = .,
    title = "Table 1",
    show.rownames = F,
    alternate.rows = T
  )
```


```{r}
skimr::skim(nhanes)
```


```{r table 1 old, eval=F, include=F}
skim(nhanes) %>%
  as_tibble() %>%
  dplyr::select(skim_variable, n_missing, numeric.mean, numeric.sd, numeric.hist) %>%
  filter(!skim_variable %in% c("SEQN", "fasting_subsample_weight")) %>% 
  mutate(numeric.mean = round(numeric.mean, 2), numeric.sd = round(numeric.sd, 2)) %>%
  rename(missing = n_missing, mean = numeric.mean, `standard deviation` = numeric.sd, distribution = numeric.hist) %>%
  gt(rowname_col = "skim_variable", caption = "Table 1")
```

```{r omit missing values}
# remove rows with missing values
# remove rows of participants whom provided blood specimen but did not meet the 8 to less than 24 hour fasting criteria
nhanes <- na.omit(nhanes) %>%
  filter(fasting_subsample_weight != 0)
```

```{r check variable distributions, fig.width=7, fig.height=4}
p1 <- nhanes %>%
  ggplot(aes(x=diet_health)) +
  geom_bar()

p2 <- nhanes %>%
  ggplot(aes(x=age)) +
  geom_histogram(bins=30)

p3 <- nhanes %>%
  ggplot(aes(x=glucose)) +
  geom_histogram(bins=50)

p4 <- nhanes %>%
  ggplot(aes(x=sleep_hrs)) +
  geom_bar()

p5 <- nhanes %>%
  ggplot(aes(x=gender)) +
  geom_bar()

p6 <- nhanes %>%
  ggplot(aes(x=bmi)) +
  geom_histogram(bins=30)

p7 <- nhanes %>%
  ggplot(aes(x=alcohol_frequency)) +
  geom_bar()

p8 <- nhanes %>%
  ggplot(aes(x=met)) +
  geom_histogram(bins=50)

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8)
```

## Univariate Analyses

```{r determine functional form of diet health}
mfp(glucose ~ fp(diet_health), data = nhanes) # fractional polynomials suggests linear encoding of diet health score

anova(
  lm(glucose ~ diet_health, data = nhanes),
  lm(glucose ~ factor(diet_health), data = nhanes)
) # categorical encoding of diet health doesn't improve model fit

# keep diet health as linear
```
```{r diet health univariate model}
lm(glucose ~ diet_health, data = nhanes) %>% summary()

nhanes %>%
  ggplot(aes(x=diet_health, y=glucose)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm")

#lm(glucose ~ diet_health, data = nhanes) %>% autoplot()
```

```{r sleep hours univariate model}
lm(glucose ~ sleep_hrs, data = nhanes) %>% summary()

nhanes %>%
  ggplot(aes(x=sleep_hrs, y=glucose)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm")

```

```{r determine functional form of MET score}
mfp(glucose ~ fp(met), data = nhanes) # fractional polynomials suggests log transformation

anova(
  lm(glucose ~ met, data = nhanes),
  nhanes %>% dplyr::mutate(met.quint = quantcut(met, q = 5)) %>% lm(glucose ~ met.quint, data = .)
) # improves model fit

# I will use a log-transformation of met score

nhanes <-
  nhanes %>%
  mutate(met_log = log(((met+4)/10000)))
```

```{r met activity score univariate model}
lm(glucose ~ met_log, data = nhanes) %>% summary()

nhanes %>%
  ggplot(aes(x=met_log, y=glucose)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm")
```

```{r determine functional form of sleep hours}
mfp(glucose ~ fp(sleep_hrs), data = nhanes) # suggests a linear encoding, I will keep sleep hours as linear
```

```{r}
lm(glucose ~ sleep_hrs, data = nhanes) %>% summary()

ggplot(data = nhanes, aes(x=sleep_hrs, y=glucose)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm")
```
```{r}
tab_model(
  lm(glucose ~ met_log, data = nhanes),
  lm(glucose ~ diet_health, data = nhanes),
  lm(glucose ~ sleep_hrs, data = nhanes),
  pred.labels = c("(Intercept)", "log(MET)", "Diet Health", "Sleep Hours"),
  CSS = list(
    css.depvarhead = 'color: red;',
    css.centeralign = 'text-align: left;', 
    css.firsttablecol = 'font-weight: bold;', 
    css.summary = 'color: blue;'
  ),
  title = "Table 2"
)
```


```{r unadjusted model}
unadj.m <- lm(glucose ~ met_log + diet_health + sleep_hrs, data = nhanes)
summary(unadj.m)
autoplot(unadj.m)


# check normality
psych::describe(unadj.m$residuals)
hist(unadj.m$residuals, breaks = 30)
shapiro.test(unadj.m$residuals)
```
Check functional form for other variables

```{r age functional form}
mfp(glucose ~ fp(age), data = nhanes) # I((age/100)^2)+I((age/100)^3)

anova(
  lm(glucose ~ met, data = nhanes),
  nhanes %>% dplyr::mutate(age.quint = quantcut(age, q = 5)) %>% lm(glucose ~ age.quint, data = .)
) # improves model fit

# use age quintile, as it is easier to interpret than a square + cubed term

nhanes <-
  nhanes %>%
  dplyr::mutate(age.quint = quantcut(age, q = 5))
```

```{r convert gender to categorical}
nhanes <-
  nhanes %>%
  mutate(gender.f = factor(gender))
```

```{r alcohol frequency functional form}
mfp(glucose ~ fp(alcohol_frequency), data = nhanes) # I(((alcohol_frequency+1)/10)^-2)+log(((alcohol_frequency+1)/10))

anova(
  lm(glucose ~ alcohol_frequency, data = nhanes),
  nhanes %>%
    mutate(alcohol_frequency.f = factor(alcohol_frequency)) %>%
    lm(glucose ~ alcohol_frequency.f, data = .)
) # better fit (p<0.001)

# convert to categorical instead of complex transformation
nhanes <-
  nhanes %>%
    mutate(alcohol_frequency.f = factor(alcohol_frequency))
```

```{r bmi functional form}
nhanes %>%
  ggplot(aes(x=bmi, y=glucose)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm", se = T)

summary(lm(glucose ~ bmi, data = nhanes))

mfp(glucose ~ fp(bmi), data = nhanes) # 1/sqrt(bmi) is the suggested transformation for bmi

summary(lm(glucose ~ I(1/sqrt(bmi)), data = nhanes))

nhanes %>%
  ggplot(aes(x=1/sqrt(bmi), y=glucose)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm", se = T)

# convert bmi to categorical
nhanes <-
  nhanes %>%
  mutate(bmi_cat = cut(bmi, breaks = c(0, 18.5, 25, 30, 100), right = F, labels = c("Underweight", "Healthy Weight", "Overweight", "Obese"))) %>%
  mutate(bmi_cat = relevel(bmi_cat, ref="Healthy Weight"))

nhanes %>%
  ggplot(aes(x=bmi_cat, y=glucose, fill=bmi_cat)) +
  geom_boxplot()

summary(lm(glucose ~ bmi_cat, data = nhanes))


## the 1/sqrt(bmi) transformation offers highest adjusted R-squared so I will use that as the encoding for BMI
nhanes <-
  nhanes %>%
  mutate(bmi_sqrt = 1/sqrt(bmi))
```


## Check interaction terms

```{r interactions}
# build full model
# test significance of interactions by removing individual interaction terms

# full model
lm(glucose ~ alcohol_frequency.f * diet_health + alcohol_frequency.f * sleep_hrs + bmi_cat * met_log , data = nhanes) %>% summary()

anova(
  lm(glucose ~ diet_health + alcohol_frequency.f * sleep_hrs + bmi_cat * met_log , data = nhanes),
  lm(glucose ~ alcohol_frequency.f * diet_health + alcohol_frequency.f * sleep_hrs + bmi_cat * met_log , data = nhanes)
) # alcohol frequency and diet health interaction not significant (p = 0.23)

anova(
  lm(glucose ~ alcohol_frequency.f * diet_health + sleep_hrs + bmi_cat * met_log , data = nhanes),
  lm(glucose ~ alcohol_frequency.f * diet_health + alcohol_frequency.f * sleep_hrs + bmi_cat * met_log , data = nhanes)
) # alcohol frequency and sleep hours interaction not significant (p = 0.37)

anova(
  lm(glucose ~ alcohol_frequency.f * diet_health + alcohol_frequency.f * sleep_hrs + bmi_sqrt + met_log , data = nhanes),
  lm(glucose ~ alcohol_frequency.f * diet_health + alcohol_frequency.f * sleep_hrs + bmi_sqrt * met_log , data = nhanes)
) # bmi and met activity score not significant (p = 0.27)
```
It looks like met activity score and bmi interact (p = 0.097) at an alpha level of 0.15, so I will keep it in the model


## Check for confounders

```{r confounding check}
# reduced output for viewing
tab_model(
  lm(glucose ~ diet_health + sleep_hrs + met_log , data = nhanes),
  lm(glucose ~ diet_health + sleep_hrs + met_log + age, data = nhanes),
  lm(glucose ~ diet_health + sleep_hrs + met_log + gender.f, data = nhanes),
  lm(glucose ~ diet_health + sleep_hrs + met_log + age + gender.f, data = nhanes),
  terms = c("diet_health", "sleep_hrs", "met_log"),
  pred.labels = c("Diet Health Score", "Sleep Hours", "log(MET)"),
  dv.labels = c("Unadjusted", "Age Adjusted", "Gender Adjusted", "Age and Gender Adjusted"),
  title = "Supplementary Table 1"
)
```

Age and gender seems to confound the relationship between MET and glucose as well as diet health and glucose. I will leave them both in the final model.

### Preliminary Final model
```{r}
m1 <- lm(glucose ~ diet_health + sleep_hrs + met_log + age + gender, data = nhanes)
summary(m1)
```

```{r}
autoplot(m1)
```
Normality and Homoscedasticity appear to be violated

```{r assess normality of residuals for preliminary final model}
ggplot() +
  geom_qq(aes(sample = rstandard(m1))) +
  geom_abline(color = "red")

ggplot(data = nhanes, aes(x = m1$residuals)) +
    geom_histogram(fill = 'steelblue', color = 'black', bins=30) +
    labs(title = 'Figure 2', subtitle = 'Histogram of Residuals', x = 'Residuals', y = 'Frequency')
```

Try transforming y to log(y+c), where c = 1 - min(y) 

```{r log transform glucose}
nhanes <- nhanes %>%
  mutate(glucose_log = log(glucose + (1 - min(glucose))))

m2 <- lm(glucose_log ~ diet_health + sleep_hrs + met_log + age + gender, data = nhanes)

autoplot(m2)
```

```{r reassess residuals}
f4 <- ggplot() +
  geom_qq(aes(sample = rstandard(m2))) +
  geom_abline(color = "red") +
  labs(title = "Figure 4", subtitle = "Normal Q-Q")

f5 <- ggplot(data = nhanes, aes(x = m1$residuals)) +
    geom_histogram(fill = 'steelblue', color = 'black', bins=30) +
    labs(title = 'Figure 5', subtitle = 'Histogram of Residuals', x = 'Residuals', y = 'Frequency')

grid.arrange(f4, f5, nrow=1)
```
Normality looks better now. Homoscedasticity looks better as well.

## Check influential points

```{r collinearity}
ols_vif_tol(m2) 
```
No evidence of collinearity

```{r cooks distance}
# Cook's Distance quantifies influence
ols_plot_cooksd_bar(m2)
```
No points with Cook's D > 0.5

```{r DFFITS}
ols_plot_dffits(m2)
```

```{r DFBETAS}
ols_plot_dfbetas(m2)

dfbetas_plots <- ols_plot_dfbetas(m2, print_plot = F)$plots

grid.arrange(
  dfbetas_plots[[1]], dfbetas_plots[[2]], dfbetas_plots[[3]],
  dfbetas_plots[[4]], dfbetas_plots[[5]], dfbetas_plots[[6]], 
  nrow = 2
)
```


```{r}
ols_plot_resid_stud(m2)
```

```{r}
ols_plot_resid_lev(m2)
```
Observations 3396, 2519, and 2460 seems to be influential points.

```{r}
cbind(
  pred = predict(m2, newdata = nhanes[c(1013, 2915, 2055, 1945, 3086, 3090, 1941, 2347),]),
  nhanes[c(1013, 2915, 2055, 1945, 3086, 3090, 1941, 2347), c(ncol(nhanes), seq_along(names(nhanes)[-ncol(nhanes)]))]
) %>%
  select(c(SEQN, pred, glucose_log, gender, age, sleep_hrs, diet_health, met_log, met)) %>%
  mutate(exp_pred = exp(pred), glucose = exp(glucose_log)) %>%
  select(-c(pred, glucose_log))

```

The data points look reasonable, there is no compelling reason to remove them from the model. These individuals had abnormal average hours of sleep and low activity scores.

Perform sensitivity analysis to see how much parameter estimates change without these points.

```{r sensitivity analysis}
lm(glucose_log ~ diet_health + sleep_hrs + met_log + age + gender, data = nhanes[-c(1013, 2915, 2055, 1945, 3086, 3090, 1941, 2347), ]) %>%
  summary()

tab_model(
  m2, lm(glucose_log ~ diet_health + sleep_hrs + met_log + age + gender, data = nhanes[-c(1013, 2915, 2055, 1945, 3086, 3090, 1941, 2347), ]),
  dv.labels = c("Original Model", "Influential Points Removed"),
  pred.labels = c("(Intercept)", "Diet Health Score", "Sleep Hours", "log(MET)", "Age", "Gender"),
  title = "Supplementary Table 2"
)
```
Estimates and standard errors are very similar, and there doesn't appear to be a data issue. I will leave these observations in the model.

```{r table 3}
tab_model(
  m2,
  pred.labels = c("(Intercept)", "Diet Health Score", "Sleep Hours - Weekdays or Weekends", "log(MET)", "Age", "Gender"),
  dv.labels = c("log(Fasting Glucose (mg/dL))"),
  CSS = list(
    css.depvarhead = 'color: red;',
    css.centeralign = 'text-align: left;', 
    css.firsttablecol = 'font-weight: bold;', 
    css.summary = 'color: blue;'
  ),
  title = "Table 3"
)
```

```{r interpreting transformations of y}
# diet health
100*(exp(0.0467504) - 1)

# sleep hours
100*(exp(-0.0028154) - 1)

100*(exp(-0.0067134) - 1)
```

